---
title: "Machine Learning Course Project"
author: "San Emmanuel James"
date: "20 August 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

The goal of this project is to apply machine learning algorithms to the 20 test cases available in the test and train accelerometer data to predict the manner in which they did the exercise. This is the "classe" variable in the training set. Where possible, use any of the other variables to predict with. Then create a report describing how model was built and cross validation was used, what the expected out of sample error is, and reasons for the choices for made. 

## Project Setup

### Reproduceability

A seed of 12345 was set to ensure reproduceability. 

Packages used include;
a) Caret v6.0.70
b) RandomForest v4.6.12 
c) Rpart v4.1.10
d) Rpart.plot v2.0.1
e) e1071 v1.6.7

Changing the versions may result in varying results but this variance should be quite minimal.

### Building the Model

The outcome variable for this project is "classe", a factor variable with 5 levels. In the assignment data, “participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

* Exactly according to the specification (Class A)
* Throwing the elbows to the front (Class B)
* Lifting the dumbbell only halfway (Class C)
* Lowering the dumbbell only halfway (Class D)
* Throwing the hips to the front (Class E)

Class A corresponds to the right way to perform the exercise, while the other 4 classes correspond to common mistakes while performing the exercise. Our prediction evaluation will thus be focused on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction.
We will test two models using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as our final model.

### Cross-validation

Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: subTraining data (75% of the original Training data set) and subTesting data (25%). The models will then be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

### Expected out-of-sample error

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

### Reasons for my choices

The outcome variable “classe” is an unordered factor variable. Thus, we can choose our error type as 1-accuracy. We have a large sample size with N= 19622 in the Training data set. This allow us to divide our Training sample into subTraining and subTesting to allow cross-validation. Features with all missing values will be discarded as well as features that are irrelevant. All other features will be kept as relevant variables.
Decision tree and random forest algorithms are known for their ability of detecting the features that are important for classification [2]. Feature selection is inherent, so it is not so necessary at the data preparation phase. Thus, there won’t be any feature selection section in this report.

### Code and Results
Packages, Libraries, Seed

Installing packages, loading libraries, and setting the seed for reproduceability:

```{r }
#if packages not yet installed;
#install.packages("caret")
#install.packages("randomForest")
#install.packages("rpart")

#Load packages
library(caret)
library(randomForest) #Random forest for classification and regression
library(rpart) # Regressive Partitioning and Regression trees
library(rpart.plot) # Decision Tree plot
library(e1071)

# set seed for reproduceability
set.seed(12345)
```

### Loading data sets and preliminary cleaning

Here we load data, code missing values and drop unnecessary variables

```{r}
# After saving both data sets into my working directory
# Some missing values are coded as string "#DIV/0!" or "" or "NA" - these will be changed to NA.
# We notice that both data sets contain columns with all missing values - these will be deleted.  
rm(list = ls()) #clear workspace

setwd("H:/Data Science/Machine Learning/week 4/Project/") #set working directory

#initialize variable with file download URL
fileURL1 = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv" #Initialize file URL for training set data download
fileURL2 = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv" #Initialize file URL for testing set data download

if(!file.exists("Data")){ #check if data directory exists, and create
      dir.create("Data")
}
if(!file.exists("Data/pml-testing.csv")){ #check if data file exists and download
      download.file(fileURL1, destfile = "Data/pml-testing.csv")
}
if(!file.exists("Data/pml-training.csv")){ #check if data file exists and download
      download.file(fileURL2, destfile = "Data/pml-training.csv")
}
# Loading the training data set into my R session replacing all missing with "NA"
trainset <- read.csv("Data/pml-testing.csv", na.strings=c("NA","#DIV/0!", ""))

# Loading the testing data set 
testset <- read.csv("Data/pml-training.csv", na.strings=c("NA","#DIV/0!", ""))

# Check dimensions for number of variables and number of observations
dim(trainset)
dim(testset)

# Drop columns with all values missing
trainset<-trainset[,colSums(is.na(trainset)) == 0]
testingset <-testset[,colSums(is.na(testset)) == 0]

# Some variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). We can delete these variables.
trainset   <-trainset[,-c(1:7)]
testset <-testset[,-c(1:7)]

```
```{r eval=FALSE}
# and have a look at our new datasets:
dim(trainset)
dim(testset)
head(trainset)
head(testset)
```

### Partition the training data set to allow cross-validation

The training data set contains 53 variables and 19622 obs.
The testing data set contains 53 variables and 20 obs.
In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (75%) and subTest (25%).
This will be performed using random subsampling without replacement.

```{r echo=FALSE}
subsamples <- createDataPartition(y=trainset$classe, p=0.75, list=FALSE)
subTraining <- trainset[subsamples, ] 
subTesting <- trainset[-subsamples, ]

```
```{r eval=FALSE}
dim(subTraining)
dim(subTesting)
head(subTraining)
head(subTesting)
```

### A look at the Data

The variable “classe” contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and compare one another.
```{r}
plot(subTraining$classe, col="gold", main="Bar Plot of levels of the variable classe within the subTraining data set", xlab="classe levels", ylab="Frequency")
```

From the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent with more than 4000 occurrences while level D is the least frequent with about 2500 occurrences.

## First prediction model: Using Decision Tree

```{r}
model1 <- rpart(classe ~ ., data=subTraining, method="class")

# Predicting:
prediction1 <- predict(model1, subTesting, type = "class")

# Plot of the Decision Tree
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)

# Test results on our subTesting data set:
confusionMatrix(prediction1, subTesting$classe)
```

## Second prediction model: Using Random Forest

```{r}
model2 <- randomForest(classe ~. , data=subTraining, method="class")

# Predicting:
prediction2 <- predict(model2, subTesting, type = "class")

# Test results on subTesting data set:
confusionMatrix(prediction2, subTesting$classe)
```

### Decision

As expected, Random Forest algorithm performed better than Decision Trees.
Accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to 0.739 (95% CI: (0.727, 0.752)) for Decision Tree model. The random Forest model is choosen. The accuracy of the model is 0.995. The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

## Submission
```{r}
# predict outcome levels on the original Testing data set using Random Forest algorithm
predictfinal <- predict(model2, testingset, type="class")
predictfinal
```

```{r}
# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)
```

## References
[1] Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human ’13) . Stuttgart, Germany: ACM SIGCHI, 2013.

[2] Krzysztof Gra'bczewski and Norbert Jankowski. Feature Selection with Decision Tree Criterion.
